Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Final State with Multivariate Techniques

### 2015-12-07 paper/HH4b.tex
more work on the paper

\abstract{
  The measurement of Higgs pair production is one of the cornerstones of
  the LHC program for the next years.
  %
  Double Higgs production provides a unique window to the electroweak
  symmetry breaking mechanism, and in particular it allows
  to access the Higgs trilinear coupling.
  %
  In this work we study the feasibility of the observation of the production of
  Higgs boson pairs at the High-Luminosity LHC (HL-LHC) in the
  $b\bar{b}b\bar{b}$ final state.
  %
  While the signal in this channel
  is enhanced by the large branching fraction of Higgs bosons to
  $b\bar{b}$ pairs, the measurement is  challenging due to the
  overwhelming QCD multijet background.
  %
  Our analysis is based on a combination of traditional cut-based
  methods with modern multivariate techniques, which lead to a substantial
  enhacement of the signal discrimination.
  %
  We account for  all relevant
  backgrounds, including the contribution from light jet fakes that is found
  to be as important or even dominant over the irreducible $4b$ QCD background.
  %
  We demonstrate the robustness of our analysis under realistic
  pile-up conditions expected at the HL-LHC.
  %
   By combining the contribution from the boosted, intermediate and resolved
  topologies we obtain a signal significance of $S/\sqrt{B}\simeq 4$.
  %
  Our analysis indicates
  that the $b\bar{b}b\bar{b}$ channel 
  should be enough claim evidence for
  the observation of double Higgs production
  at the HL-LHC.
}

>>>>

\abstract{
  The measurement of Higgs pair production is one of the cornerstones of
  the LHC program for the next years.
  %
  Double Higgs production provides a unique window to the electroweak
  symmetry breaking mechanism due to its sensitivity
  to the Higgs trilinear coupling.
  %
  In this work, we study the feasibility of a measurement of
  Higgs pair production in the
  $b\bar{b}b\bar{b}$ final state at the LHC.
  %
  Our analysis is based on a combination of traditional cut-based
  methods with state-of-the-art
  multivariate techniques.
  %
  We account for  all relevant
  backgrounds, including the contribution from light and charm
  jet mistags that is found to be comparable
  to the irreducible $4b$ QCD background.
  %
  We demonstrate the robustness of our analysis in a high pile-up
  environment.
  %
  For an integrated luminosity of $\mathcal{L}=3$ ab$^{-1}$,
  an overall signal significance of  $S/\sqrt{B}\simeq 4$ is obtained.
  %
  Our analysis indicates
  that the $b\bar{b}b\bar{b}$ channel
alone should allow the observation of double Higgs production
  at the High Luminosity LHC.
}

### 2015-12-07 paper/sec-analysis.tex
more work on the paper

One of the selection cuts with highest discrimination power is the requirement
that the invariant mass of the Higgs candidate (di)jets must lie within a window
around the nominal Higgs value.
%
In Fig.~\ref{fig:mHHinv} we show the invariant mass
of the leading reconstructed Higgs candidates, before the Higgs mass window
cut Eq.~(\ref{higgsmasswindow})
  is applied, both for the resolved and boosted categories.
%
The signal distribution is of course peaked at the
nominal Higgs mass of $m_h=125$ GeV.
%
In the boosted category, the background shows no particular
structure, while remarkable in the resolved category the background
also shows a peak-like structure, artificially induced by the
acceptance kinematical cuts on the jets.
%
Note that the applied smearing, which mimics detector resolution effects,
leads to a rather broad distribution which reduces the usefulness
of the Higgs mass window cut.
%
While our Higgs mass window cut is relatively loose,
information of the different shape of the $m_{h}$
distribution will still be exploited by the MVA.
%

@@

Another important kinematical distribution of this process is the invariant mass
of the di-Higgs system.
%
First of all because this is a direct measure of the boost of the system,
and also because in many BSM scenarios this distribution can be substantially
modified as compared to the SM case, for example in the presence
of certain dimension-6 EFT operators~\cite{Azatov:2015oxa}.
%
Indeed, one important advantage of the $4b$ final state for
di-Higgs production is that it significantly increases the reach
in $m_{hh}$ as compared to other channels with smaller branching
ration,
such as $2b2\gamma$.
%
With this motivation, we show in
Fig.~\ref{fig:mhh} the invariant mass distribution of the
reconstructed Higgs pairs,
comparing the resolved category (left) with the boosted category (right).

In the resolved case, we see that the distribution
in $m_{hh}$ is rather harder for the signal as for the background,
and thus one expects that cutting in $m_{hh}$ would help signal
discrimination: we will verify this with the help of the MVA.
%
For the boosted category the trend of the $m_{hh}$ distribution
is different because of the jet selection cuts, with the
distribution now peaking at higher values.
%
In this case signal and background distributions
look reasonably similar.
%
Note that at parton-level the $m_{hh}$ distribution as a kinematical
cut-off at $m_{hh}^{\rm min}=250$ GeV, which is modified by both
shower effects and by detector resolution effects.
%
While we don not cut in this quantity, it is
one of the inputs for the MVA to help to improve signal discrimination.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.48\textwidth]{plots/m_HH_C2_res_noPU.pdf}
  \includegraphics[width=0.48\textwidth]{plots/m_HH_C2_bst_noPU.pdf}
  \caption{\small Left plot: the invariant mass distribution of the Higgs
    pair candidates, $m_{hh}$, comparing signal and background events,
    in the resolved category.
    %
    Right plot: same for the boosted category.
}
\label{fig:mhh}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

Another interesting distribution is the transverse momentum of
the di-Higgs system.
%
In Fig.~\ref{fig:pthh} we show the $p_T^{hh}$
distribution
for the resolved and boosted categories.
%
Again we see that the background has a steeper $p_T^{hh}$ distribution
that the signal, in both categories, thus this variable
should provide some additional discrimination power, and therefore
we will use it as another of the MVA inputs.
%
Note that in our LO simulation this distribution is generated purely
by the parton shower - a more refined calculation would require
either the matching with higher-multiplicity matrix elements~\cite{Maierhofer:2013sha} or
the full NLO calculation~\cite{Frederix:2014hta}, to treat properly the first hard emission.
%
Nevertheless, the MVA shows only limited sensitivity to this variable, so its
modelling appears not to be crucial in our case.

@@

\subsection{PU subtraction with {\tt SoftKiller}}
\label{sec:pileup}

To study the impact of PU in our analysis,
Minimum Bias (MB) events,
including Multiple Parton Interactions (MPI), have been generated
with {\tt Pythia8}, and
superimposed to the signal
and background samples described in Sect.~\ref{mcgeneration}.
%
We have explored two scenarios for the impact
of PU in our analysis,
one with a mean number of
PU vertices per bunch crossing of $\la n_{\rm PU}\ra=80$, and another
with $\la n_{\rm PU}\ra=150$.
%
In order to subtract the PU, a number of techniques
have been developed
recently~\cite{Cacciari:2009dp,TheATLAScollaboration:2013pia,Butterworth:2008iy,Cacciari:2007fd,Krohn:2009th,Krohn:2013lba,Ellis:2009me,Bertolini:2014bba,Cacciari:2014gra,Cacciari:2014jta,Berta:2014eza,Larkoski:2014wba}.\footnote{
These techniques have also important applications in the subtraction
of the UE/MPI contamination for jet reconstruction
in heavy ion collisions~\cite{Cacciari:2010te}.
}
%
In this work, PU  will be subtracted by means
of the the {\tt SoftKiller} (SK)
method~\cite{Cacciari:2014gra}, as implemented in {\tt FastJet}.
%

The idea underlying {\tt SoftKiller} is based on eliminating particles
below a given cut-off in their transverse momentum, $p_T^{\rm (cut)}$, whose
value is dynamically determined in a way that makes the event-wide
transverse-momentum flow density $\rho$ vanish.
%
This $p_T$ flow density is defined as
\be
\rho\equiv{\rm median}_i \Bigg\{ \frac{p_{Ti}}{A_i}\Bigg\} \, ,
\ee
where the median is computed over all the patches $i$ with area
$A_i$ and transverse momentum $p_{Ti}$ in which the $\lp \eta,\phi\rp$ plane
is partitioned.
%
From its definition in terms of the median,
we observe that the value of $p_T^{(\rm cut)}$
will be dynamically raised until half of the patches have $\rho=0$.
%
The size and number of these patches is a free parameter of the algorithm -
here we will use square patches with length $a=0.4$.
%
We restrict ourselves to the central rapidity region,
$|\eta| \le 2.5$, for the estimation of the
$p_T$ flow density $\rho$.
%
In this analysis, {\tt SoftKiller} method is applied
to particles at the end of the parton shower, before
jet clustering.

To validate PU subtraction,
we now compare different kinematical distributions
in the case without PU and in the case
with PU subtracted with {\tt SoftKiller}.
%
Then we will also compare some specific distributions
also in the case where PU has not been subtracted.
%
We find that
{\tt SoftKiller} exhibits a reasonable
performance for the PU subtraction.
%
We  also obtain that the boosted category is less affected
by PU than the resolved category, as expected from the higher
$p_T$ threshold in the corresponding selection cuts.

First of all, let us validate both the implementation of our
simulation of PU, as well as the SK subtraction, by comparing the
invariant mass distributions of Higgs candidates in signal
events in the resolved category.
%
In Fig.~\ref{fig:PUvalidation} we show three curves: without PU,
with PU $\la n_{PU}\ra=80$ but without any subtraction, and the
same but now with the SK subtraction.
%
As expected, if PU is not subtracted there is a large shift in the Higgs mass
peak, by about 40 GeV.
%
Once SK subtraction is performed, we recover a distribution much closer
to the original ones, with only a small shift of $\simeq 5$ GeV in the mass
distribution.
%
As we will see below, the performance of SK is even better in the boosted category,
since the effects of PU are mitigated for high $p_T$ jets.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \begin{center}
  \includegraphics[width=0.68\textwidth]{plots/m_htot_res_signal_PUnoSK.pdf}
    \caption{\small
    The invariant mass distributions of Higgs candidates in signal
events, in the resolved category, comparing the results without PU,
with PU $\la n_{PU}\ra=80$ but without any subtraction, and the
corresponding results now with SK subtraction.
}
\label{fig:PUvalidation}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

Next, we compare more
 signal distributions with
and without PU, with $\la n_{\rm PU}\ra=80$ in the
former
case (and SK subtraction).
%
In Fig.~\ref{fig:m_H_PU} we show the invariant mass distribution
of the leading and subleading Higgs candidates, corresponding to both
the boosted
and resolved categories.
%
These distributions are plotted after the $b$-tagging, that is,
before they are used as input to the MVA.
%
As we can see, in the boosted category, the residual effects of PU
after the {\tt SoftKiller} subtraction are rather mild,
with the position of the Higgs mass peaks essentially
unchanged, and only a moderate distortion of the
distribution found.
%
The effects are more important for the resolved category, where PU
shifts the Higgs peak by an amount $\Delta m_h \simeq 5$ GeV.

@@

Now we perform a similar comparison this time for
the resolved category.
%
In Fig.~\ref{fig:signal-vs-back-resolved} we compare
the kinematical distributions for signal and background events,
     with and without PU, for the invariant mass and the $p_T$ of the leading
     Higgs candidate.
     %
     Also in this
     case the PU-subtracted background distributions appear reasonably close
     to their no PU counterparts.
     %
     Therefore, these results
     suggest that the broad pattern of the signal over background
     discrimination provided by the MVA in Sect.~\ref{sec:mva},
     based on the corresponding differences between kinematical
     distributions,
will
     be maintained in the presence of PU.
     %
     In the next section we verify this expectation.

@@

>>>>

One of the selection cuts with highest discrimination power is the requirement
that the invariant mass of the Higgs candidate (di)jets must lie within a window
around the nominal Higgs value.
%
In Fig.~\ref{fig:mHHinv} we show the invariant mass
of the leading reconstructed Higgs candidates, before the Higgs mass window
cut Eq.~(\ref{higgsmasswindow})
  is applied, for the resolved and boosted categories.
%
While the signal distribution is of course peaked at the
nominal Higgs mass, the background distributions
show no particular
structure.
%
As discussed in Sect.~\ref{sec:detectormodeling}, the
width of the Higgs mass peak arises both from QCD effects
like ISR and out-of-cone radiation, as well
as from the smearing applied to final state particles.
%
While the Higgs mass window requirements
in the cut-based analysis are relatively loose, the
information of the different shape of the $m_{h}$
distribution will still be fully exploited by the MVA.
%

@@

Another important kinematical distribution of this process is the invariant mass
of the di-Higgs system.
%
This is a direct measure of the boost of the system,
which in  BSM scenarios can be substantially
enhanced, for example due to
of certain $d=6$ EFT operators~\cite{Azatov:2015oxa}.
%
One important advantage of the $b\bar{b}b\bar{b}$ final state for
di-Higgs production is that it significantly increases the reach
in $m_{hh}$ as compared to other channels with smaller branching
ratios,
such as $2b2\gamma$ or $2b2\tau$.
%
With this motivation, we show in
Fig.~\ref{fig:mhh} the invariant mass distribution of the
reconstructed Higgs pairs,
comparing the resolved and the boosted categories.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.48\textwidth]{plots/m_HH_C2_res_noPU.pdf}
  \includegraphics[width=0.48\textwidth]{plots/m_HH_C2_bst_noPU.pdf}
  \caption{\small
Same as   Fig.~\ref{fig:cutplots1} for the invariant
mass distribution of the di-Higgs system $m_{hh}$, in
the resolved (left plot) and boosted (right plot) categories.
}
\label{fig:mhh}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the resolved case, we see that the distribution
in $m_{hh}$ is rather harder for the signal as compared
to the background,
and thus one expects that cutting in $m_{hh}$ would help signal
discrimination: we will verify this with the MVA.
%
For the boosted category the overall trend of the $m_{hh}$ distribution
is different because of the jet selection cuts, with the
distribution now peaking at higher values of the invariant mass.
%
In this case signal and background distributions
look reasonably similar.
%
Note that at parton-level the $m_{hh}$ distribution
for signal events has a kinematical
cut-off at $m_{hh}^{\rm min}=250$ GeV, which is modified due to both
the parton shower and to detector resolution effects.
%

Finally, in Fig.~\ref{fig:pthh} we show the transverse momentum of
the di-Higgs system, $p_T^{hh}$,
for the resolved and boosted categories.
%
Again we see that the background has a steeper $p_T^{hh}$ distribution
that the signal, in both categories, thus this variable
should provide additional discrimination power, and therefore
we will use it as another of the MVA inputs.
%
Note that in our LO simulation the $p_T^{hh}$ distribution is generated
purely
by the parton shower - a more refined treatment would require
either the matching with higher-multiplicity
matrix elements~\cite{Maierhofer:2013sha} or
the full NLO calculation~\cite{Frederix:2014hta},
%
Nevertheless, as we will show,
the MVA shows only limited sensitivity to this variable, so its
modelling appears not to be crucial in our case.

@@

\subsection{The impact of pile-up}
\label{sec:pileup}

The results shown so far did not accounted for the
effects of PU.
%
To study the impact of PU in our analysis,
Minimum Bias (MB) events,
including Multiple Parton Interactions (MPI), have been generated
with {\tt Pythia8}, and
superimposed to the signal
and background samples described in Sect.~\ref{mcgeneration}.
%
We have explored two scenarios for the impact
of PU in our analysis,
one with a mean number of
PU vertices per bunch crossing of $\la n_{\rm PU}\ra=80$, and another
with $\la n_{\rm PU}\ra=150$.
%
While the latter is closer to the HL-LHC forecasts, since it
is beyond the scope of this paper to fully optimize
PU subtraction, we will adopt $\la n_{\rm PU}\ra=80$ as our baseline.

%
In order to subtract the PU, a number of techniques
have been developed
recently~\cite{Cacciari:2009dp,TheATLAScollaboration:2013pia,Butterworth:2008iy,Cacciari:2007fd,Krohn:2009th,Krohn:2013lba,Ellis:2009me,Bertolini:2014bba,Cacciari:2014gra,Cacciari:2014jta,Berta:2014eza,Larkoski:2014wba}.\footnote{
These techniques have also important applications in the subtraction
of the UE/MPI contamination for jet reconstruction
in heavy ion collisions~\cite{Cacciari:2010te}.
}
%
In this work, PU  will be subtracted by means
of the {\tt SoftKiller} (SK)
method~\cite{Cacciari:2014gra}, as implemented in {\tt FastJet}.
%
The idea underlying {\tt SoftKiller} is based on eliminating particles
below a given cut-off in their transverse momentum, $p_T^{\rm (cut)}$, whose
value is dynamically determined in a way that makes the event-wide
transverse-momentum flow density $\rho$ vanish.
%
This $p_T$ flow density is defined as
\be
\rho\equiv{\rm median}_i \Bigg\{ \frac{p_{Ti}}{A_i}\Bigg\} \, ,
\ee
where the median is computed over all the patches $i$ with area
$A_i$ and transverse momentum $p_{Ti}$ in which the $\lp \eta,\phi\rp$ plane
is partitioned.
%
From its definition in terms of the median,
we observe that the value of $p_T^{(\rm cut)}$
will be dynamically raised until half of the patches have $\rho=0$.
%
The size and number of these patches is a free parameter of the algorithm -
here we will use square patches with length $a=0.4$.
%
We restrict ourselves to the central rapidity region,
$|\eta| \le 2.5$, for the estimation of the
$p_T$ flow density $\rho$.
%
In this analysis, {\tt SoftKiller} method is applied
to particles at the end of the parton shower, before
jet clustering.

To validate PU subtraction,
we now compare different kinematical distributions
in the case without PU and in the case
with PU subtracted with {\tt SoftKiller}.
%
We will find that
{\tt SoftKiller} exhibits a good
performance for the PU subtraction in the present case.
%
We  also obtain that the boosted category is less affected
by PU than the resolved category, as expected from the higher
$p_T$ threshold in the corresponding selection cuts.

First of all, we compare
in Fig.~\ref{fig:PUvalidation} the
invariant mass distributions of Higgs candidates for signal
events in the resolved category.
%
We show three curves: without PU,
with PU $\la n_{\rm PU}\ra=80$ but without any subtraction, and the
same but now with the SK subtraction.
%
As expected, if PU is not subtracted there is a large shift in the Higgs mass
peak, by about 40 GeV.
%
Once SK subtraction is performed, we recover a distribution much closer
to the original ones, with only a small shift of $\simeq 5$ GeV in the mass
distribution.
%
The performance of SK is even better in the boosted category,
since the effects of PU are mitigated for high $p_T$ jets.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \begin{center}
  \includegraphics[width=0.68\textwidth]{plots/m_htot_res_signal_PUnoSK.pdf}
    \caption{\small
    The invariant mass distributions of Higgs candidates in signal
events, in the resolved category, comparing the results without PU,
with PU $\la n_{\rm PU}\ra=80$ but without any subtraction, and the
corresponding results now with SK subtraction.
}
\label{fig:PUvalidation}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

Next, we compare additional
 signal distributions with
and without PU, with $\la n_{\rm PU}\ra=80$ in the
former
case (and SK subtraction).
%
In Fig.~\ref{fig:m_H_PU} we show the invariant mass distribution
of the leading and subleading Higgs candidates, corresponding to both
the boosted
and resolved categories.
%
These distributions are plotted after the $b$-tagging, that is,
before they are used as input to the MVA.
%
As we can see, in the boosted category, the residual effects of PU
after the {\tt SoftKiller} subtraction are rather moderate,
with the position of the Higgs mass peaks essentially
unchanged, and only a moderate distortion of the
distribution found.
%
The effects are more important for the resolved category, where PU
shifts the Higgs peak by an amount $\Delta m_h \simeq 5$ GeV.

@@

Now we perform a similar comparison this time for
the resolved category.
%
In Fig.~\ref{fig:signal-vs-back-resolved} we compare
the kinematical distributions for signal and background events,
     with and without PU, for the invariant mass and the $p_T$ of the leading
     Higgs candidate.
     %
     Also in this
     case the PU-subtracted background distributions appear reasonably close
     to their no PU counterparts.

@@

These results validate our strategy for PU subtraction and indicate
that, while there is some distortion in the shape of kinematical
distributions as compared to the case without PU, it is certainly
not enough to wash out the most important differences between
signal and background events.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
